---
title: "Code Chunks"
author: "Ntemena Kapula"
date: "12/22/2020"
output: 
  html_document:
    theme: "cosmo"
---

This R markdown consists of three chunks of code.

You can quickly insert chunks of code by using 3 simple shortcuts 

1. cmd + Option + I on a mac 

2. Ctrl + Alt + I on windows

3. add chunk with the "+c insert" command in the editor toolbar (simplest)

The long way of inserting a chunk of code is by typing the chunk delimiters ```{r} and ``` (not re)

### Chunk Options

Chunk output can be customized with knitr options, arguments set in the {} of the chunk header.

Arguments:

1. __include = FALSE__: hides code and results when knit into final file. Rmarkdown still runs the code in the chunk, and the results can be used by other chunks

2. __echo = FALSE__: only displays result and not code when knit into final file. This can be a useful way to embed figures

3. __message = FALSE__: prevents messages generated by code from appearing in the finished file when knit

4. __warning = FALSE__: prevents warnings that are generated by code from appearing in the finished file

5. __fig.cap = "..."__: adds a caption to graphical results

6. __error = FALSE__: error messages generated by code are not displayed in final file when knit.

7. __prompt = FALSE__: the symbol > will be added to the start of each line of code displayed in the final file when knit

```{r prompt=TRUE}
names(knitr::knit_engines$get())
```

for more knitr options go to [Rmarkdown_Guide](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf?_ga=2.59981291.799592073.1608699192-1437317260.1608699192) and [Knitr_options](https://yihui.org/knitr/options/)

knitr will evaluate the chunk in the named language. This code shows which other languages other than `r` are supported and for which knitr will evaluate.
```{r}
names(knitr::knit_engines$get())
```

## Global Options

To set global options that apply to every chunk in your file, call knitr::opts_chunk$set in a code chunk. 




